Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2025-08-12 src/docs_paper-jtb/paper.Rmd

Let $\z \in \mathcal{Q}(m, k)$ be an $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule.
In particular, $\mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ where
\begin{equation}
\mathcal{Q}(1, k) = \{z: H_k(z) = 0\} \\
\end{equation}
are the zeroes of the $k$th Hermite polynomial $H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2)$.
The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, and $\phi(\cdot)$ is a standard Gaussian density.
GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
When $k = 1$ Equation \@ref(eq:aghq) corresponds to a Laplace approximation.
The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
Two alternatives for the matrix decomposition [@jackel2005note] are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
This estimate may be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \@ref(eq:inla1)) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points which quickly becomes intractable as $m$ increases for $k > 1$.
We propose to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
Then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ as a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
We describe this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
\begin{equation}
\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
\end{equation}
where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
Panel C of Figure \@ref(fig:aghq) illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
Inference for the latent field follows analogously to Equation \@ref(eq:nest).

>>>>

Let $\z \in \mathcal{Q}(m, k)$ be an $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule.
In particular, $\mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ where
\begin{equation}
\mathcal{Q}(1, k) = \{z: H_k(z) = 0\} \\
\end{equation}
are the zeroes of the $k$th Hermite polynomial $H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2)$.
The corresponding weighting function $\omega_k: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega_k(\z) = \prod_{j = 1}^m \omega_k(z_j)$ where $\omega_k(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, and $\phi(\cdot)$ is a standard Gaussian density.
GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega_k(\z). \label{eq:aghq}
\end{equation}
When $k = 1$ Equation \@ref(eq:aghq) corresponds to a Laplace approximation.
The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
Two alternatives for the matrix decomposition [@jackel2005note] are the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
This estimate may be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \@ref(eq:inla1)) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega_k(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega_k(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points which quickly becomes intractable as $m$ increases for $k > 1$.
We propose to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
Then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ as a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
We describe this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
\begin{equation}
\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \widetilde{\omega}(\z),
\end{equation}
where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\widetilde{\omega}(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
Panel C of Figure \@ref(fig:aghq) illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
Inference for the latent field follows analogously to Equation \@ref(eq:nest).
